From Reversible Programs to \\ Univalent Universes and Back

### 2017-07-04 upi.tex
tweak syntax of Π₂
sec 1 and 2

Starting from the physical principle of ``conservation of
information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative}, James and
Sabry~\cite{James:2012:IE:2103656.2103667} propose a family of programming
languages $\Pi$ in which computation preserve information. Technically,
computations are \emph{type isomorphisms} which, at least in the case of finite
types, clearly preserve entropy in the information-theoretic
sense~\cite{James:2012:IE:2103656.2103667}. We will identify a ``featherweight''
version of $\Pi$ to use in our formal development.  But first we illustrate the
general flavor of the family of languages with some examples.

@@

The examples below assume a representation of the type of booleans
$\bt$ as the disjoint union $\ot \oplus \ot$ with the left injection
representing $\mathsf{false}$ and the right injection representing
$\mathsf{true}$. Given an arbitrary reversible function
$\AgdaFunction{f}$ of type $a \leftrightarrow a$, we can build the
reversible function $\AgdaFunction{controlled}~\AgdaFunction{f}$ that
takes a pair of type $\bt \otimes a$ and checks the incoming boolean;
if it is false (i.e., we are in the left injection), the function
behaves like the identity; otherwise the function applies
$\AgdaFunction{f}$ to the second argument. In either case, the
incoming boolean is reconstituted to maintain reversibility.

{\small
\[\def\arraystretch{1.2}\begin{array}{rcll}
\AgdaFunction{controlled}  &:& \forall a.~ (a \leftrightarrow a) \rightarrow
                            & (\bt \otimes a \leftrightarrow \bt \otimes a) \\
\AgdaFunction{controlled}~\AgdaFunction{f} &=&

@@

The $\AgdaFunction{not}$ function is a simple lifting of
\AgdaFunction{swap₊}. Using the
\AgdaFunction{controlled} building block, we can build a
controlled-not $\AgdaFunction{cnot}$ gate and a
controlled-controlled-not gate, also known as the
\AgdaFunction{toffoli} gate. The latter gate is a universal function
for combinational boolean circuits thus showing the expressiveness of
the language.
{\small
\[\begin{array}{rcl}
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=&
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
\end{array}\]}
%%%

While we wrote \AgdaFunction{controlled} in equation-reasoning style,
\AgdaFunction{not} is written in combinator style.  These are
equivalent as $\byiso{-}$ is defined in terms of $\odot$.

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the question of when two programs are ``equivalent.'' Consider
the following six programs of type~$\bt \leftrightarrow \bt$

@@

\noindent We neither expect, nor desire, that all these programs of
the same type to be equivalent. In the above, each group of
three does represent ``the same'' program but the two groups
represent distinct programs. Thinking extensionally, i.e., by looking
at all possible input-output pairs, it is indeed easy to verify this
fact, but we can do better. We can provide \emph{evidence} (i.e., a
program that manipulates programs) that can constructively transform
every program to an equivalent one. We show such a level-2 program
proving that $\AgdaFunction{not₃}$ is equivalent to
$\AgdaFunction{not}$. For illustration, the program for
$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
the reader to map the steps below to manipulations on the diagram that
would incrementally simplify it:

@@

It is worthwhile mentioning that the above derivation could also be drawn as
one (large!) commutative diagram in an appropriate category, with each
$\byisotwo{-}$ as a $2$-arrow (and representing a natural isomorphism).
See~\cite{Shulmann} for that interpretation.

@@

\jacques{the code above uses $\odot$ for 1-composition,
$\boxdot$ for parallel 2-composition of $\odot$, while the
code below uses $\circ$ and $\odot$ respectively, which is
quite confusing.  We should pick one notation.}
\[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
(\textit{Types}) & \tau &::=& \bt \\
\\
(\textit{Terms}) &  v &::=& \fc &:& \bt \\
              && \alt & \tc &:& \bt \\
\\
 (\textit{1-combinators}) &  c &::=& \id &:& \tau \iso \tau \\
               && \alt & \swap &:& \bt \iso \bt \\
               && \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
               && \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)  \\
\\
(\textit{2-combinators}) & \alpha &::=& \id &:& c \isotwo c \\
            && \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
            && \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
            && \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
            && \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
            && \alt & \rho &:& \swap \circ \swap \isotwo \id \\
            && \alt & \assocc &:&
                                  \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
            && \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to
                             (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
            && \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
            && \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
\end{array}\]

The syntactic category $c$ is that of 1-combinators denoting
reversible programs, type isomorphisms, permutations, or equivalences
depending on one's favorite interpretation. There are two primitive
combinators $\id$ and $\swap$ which are closed under inverses $!$ and
sequential composition $\circ$. The syntactic category $\alpha$ is
that of 2-combinators denoting reversible program transformations,
coherence conditions on type isomorphisms, equivalences between
permutations, or program optimizations depending on one's favorite
interpretation.

It is relatively easy to think of a few sound program transformations
to include in the category $\alpha$ of 2-combinators. The set above
however is \emph{complete}. We also have canonical forms (constructively).

\begin{lemma}[Canonical Forms]
  Given a 1-combinator $c : \tau \leftrightarrow \tau$, we either have
  a normalization $c \Leftrightarrow \id$ or a normalization
  $c \Leftrightarrow \swap$.
\end{lemma}

@@

>>>>

Starting from the physical principle of ``conservation of
information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative}, James and
Sabry~\cite{James:2012:IE:2103656.2103667} propose a family of programming
languages $\Pi$ in which computation preserve information. Technically,
computations are \emph{type isomorphisms} which, at least in the case of finite
types, clearly preserve entropy in the information-theoretic
sense~\cite{James:2012:IE:2103656.2103667}. We illustrate the general flavor of
the family of languages with some examples and then identify a ``featherweight''
version of $\Pi$ to use in our formal development.

@@

The examples below assume a representation of the type of booleans $\bt$ as the
disjoint union $\ot \oplus \ot$ with the left injection representing
$\mathsf{false}$ and the right injection representing $\mathsf{true}$. Given an
arbitrary reversible function $\AgdaFunction{f}$ of type $a \leftrightarrow a$,
we can build the reversible function
$\AgdaFunction{controlled}~\AgdaFunction{f}$ that takes a pair of type
$\bt \otimes a$ and checks the incoming boolean; if it is false (i.e., we are in
the left injection), the function behaves like the identity; otherwise the
function applies $\AgdaFunction{f}$ to the second argument. The incoming boolean
is then reconstituted to maintain reversibility:

{\small
\[\def\arraystretch{1.2}\begin{array}{rcll}
\AgdaFunction{controlled}  &:& \forall a.~ (a \leftrightarrow a) \quad\rightarrow
                            & ~(\bt \otimes a \leftrightarrow \bt \otimes a) \\
\AgdaFunction{controlled}~\AgdaFunction{f} &=&

@@

\noindent The left column shows the sequence of types that are visited during
the computation; the right column shows the names of the combinators\footnote{We
  use names that are hopefully quite mnemonic; for the precise definitions of
  the combinators see the $\Pi$-papers.} that witness the corresponding type
isomorphism. The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware circuit) for
an automorphism on $\bt \otimes a$: it can be read top-down or bottom-up to go
back and forth.

The $\AgdaFunction{not}$ function below is a simple lifting of
\AgdaFunction{swap₊} which swaps the left and right injections in a sum
type. Using the \AgdaFunction{controlled} building block, we can build a
controlled-not ($\AgdaFunction{cnot}$) gate and a controlled-controlled-not
gate, also known as the \AgdaFunction{toffoli} gate. The latter gate is a
universal function for combinational boolean circuits thus showing the
expressiveness of the language:

{\small
\[\begin{array}{rcl}
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=&
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
\end{array}\]}
%%%

\noindent While we wrote \AgdaFunction{controlled} in equational-reasoning
style, \AgdaFunction{not} is written in the point-free combinator style.  These
are equivalent as $\byiso{-}$ is defined in terms of the sequential composition
combinator $\odot$.

As is customary in any semantic perspective on programming languages, we are
interested in the question of when two programs are ``equivalent.'' Consider the
following six programs of type~$\bt \leftrightarrow \bt$:

@@

\noindent The programs are all of the same type but this is clearly not a
sufficient condition for ``equivalence.'' Thinking extensionally, i.e., by
looking at all possible input-output pairs, it is easy to verify that the six
programs split into two classes: one equivalent to the identity function and one
equivalent to the boolean negation. In the context of $\Pi$, we can better: we
can provide \emph{evidence} (i.e., a program that manipulates programs) that can
constructively transform every program to an equivalent one. We show such a
level-2 program proving that $\AgdaFunction{not₃}$ is equivalent to
$\AgdaFunction{not}$. For illustration, the program for $\AgdaFunction{not₃}$ is
depicted in Fig.~\ref{fig:not}. We encourage the reader to map the steps below
to manipulations on the diagram that would incrementally simplify it:

@@

\noindent It is worthwhile mentioning that the above derivation could also be
drawn as one (large!) commutative diagram in an appropriate category, with each
$\byisotwo{-}$ as a $2$-arrow (and representing a natural isomorphism).
See~\cite{Shulmann} for that interpretation.

@@

% \jacques{the code above uses $\odot$ for 1-composition,
% $\boxdot$ for parallel 2-composition of $\odot$, while the
% code below uses $\circ$ and $\odot$ respectively, which is
% quite confusing.  We should pick one notation.}

\[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
(\textit{Types}) & \tau &::=& \bt \\
\\
(\textit{Terms}) &  v &::=& \fc &:& \bt \\
              && \alt & \tc &:& \bt \\
\\
 (\textit{1-combinators}) &  c &::=& \id &:& \tau \iso \tau \\
               && \alt & \swap &:& \bt \iso \bt \\
               && \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
               && \alt & \odot &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)  \\
\\
(\textit{2-combinators}) & \alpha &::=& \id &:& c \isotwo c \\
            && \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
            && \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
            && \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
            && \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
            && \alt & \rho &:& \swap \circ \swap \isotwo \id \\
            && \alt & \assocc &:&
                                  \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
            && \alt & \boxdot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to
                             (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
            && \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
            && \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
\end{array}\]

The syntactic category $c$ is that of 1-combinators denoting
reversible programs, type isomorphisms, permutations, or equivalences
depending on one's favorite interpretation. There are two primitive
combinators $\id$ and $\swap$ which are closed under inverses $!$ and
sequential composition $\odot$. The syntactic category $\alpha$ is
that of 2-combinators denoting reversible program transformations,
coherence conditions on type isomorphisms, equivalences between
permutations, or program optimizations depending on one's favorite
interpretation.

It is relatively easy to think of a few sound program transformations to include
in the category $\alpha$ of 2-combinators. But, as the following lemma
establishes, the above set is \emph{complete}:

\begin{lemma}[Canonical Forms]
  Given a 1-combinator $c : \tau \leftrightarrow \tau$, we either have a
  2-combinator of type $c \Leftrightarrow \id$ or a 2-combinator of type
  $c \Leftrightarrow \swap$. In other words, every 1-combinator has a canonical
  representation as either $\id$ or $\swap$ and the set of 2-combinators is rich
  enough to normalize $c$ to its canonical representation.
\end{lemma}

@@

{\small
\[\def\arraystretch{1.2}\begin{array}{rcll}
\AgdaFunction{$\dbracket{ \_ }_1$}  &:& (\bt \leftrightarrow \bt) \rightarrow `\bt \equiv `\bt \\
\AgdaFunction{$\dbracket{ \id }_1$} &=& `\mathbf{id}\\
\AgdaFunction{$\dbracket{ \swap }_1$} &=& `\mathbf{not}\\
\AgdaFunction{$\dbracket{ ! p }_1$} &=& (\dbracket{ p }_1)^{-1} \\
\AgdaFunction{$\dbracket{ p \odot q }_1$} &=& \dbracket{ p }_1 \circ \dbracket{ p }_1\\
\end{array}
\]}
